Detecting AI-Generated Images using ResNet

### 2022-12-14 paper/paper.tex
Update paper

We split around 20\% of the Indoor Scene Recognition Dataset with its generated pictures (using Stable Diffusion) for the validation part. The validation data is not used in the training process.

@@

\begin{table*}
  \centering
  \begin{tabular}{c|c||c|c|c|c}
    \hline
    \textbf{Dataset}      & \textbf{Model} & \textbf{Accuracy} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} \\
    \hline
    Indoor + SD (train)   & ResNet34       & {98.99\%}         & {99.07\%}          & {98.72\%}       & {99.00\%}         \\
    Indoor + SD (train)   & ResNet50       & {98.51\%}         & {98.90\%}          & {98.10\%}       & {98.50\%}         \\
    \hline
    Weather + SD          & ResNet34       & {90.63\%}         & {88.05\%}          & {94.02\%}       & {90.93\%}         \\
    Weather + SD          & ResNet50       & {95.61\%}         & {95.94\%}          & {95.25\%}       & {95.59\%}         \\
    Comic + SD            & ResNet34       & {92.87\%}         & {88.58\%}          & {98.44\%}       & {93.25\%}         \\
    Comic + SD            & ResNet50       & {98.34\%}         & {100.00\%}         & {96.68\%}       & {98.31\%}         \\
    Small Indoor + DALLE2 & ResNet34       & {51.04\%}         & {80.95\%}          & {2.72\%}        & {5.27\%}          \\
    Small Indoor + DALLE2 & ResNet50       & {51.84\%}         & {82.86\%}          & {4.65\%}        & {8.80\%}          \\
    Weather + DALLE2      & ResNet34       & {50.83\%}         & {53.33\%}          & {13.33\%}       & {21.33\%}         \\
    Weather + DALLE2      & ResNet50       & {50.67\%}         & {56.67\%}          & {5.67\%}        & {10.30\%}         \\
    \hline
  \end{tabular}
  \caption{Precision, Recall and F1-Score on the \% datasets}
  \label{tab:results-on-datasets}
\end{table*}

\section{Conclusion}

Conclusion goes here.

\section{Discussion and Future Work}

Discussion and future work go here.

>>>>

We split around 20\% of the Indoor Scene Recognition Dataset with its generated pictures (using Stable Diffusion) for the validation part. The validation data is not used in the training process.
\begin{table*}
  \centering
  \begin{tabular}{c|c||c|c|c|c}
    \hline
    \textbf{Dataset}      & \textbf{Model} & \textbf{Accuracy} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} \\
    \hline
    Indoor + SD (train)   & ResNet34       & {98.99\%}         & {99.07\%}          & {98.72\%}       & {99.00\%}         \\
    Indoor + SD (train)   & ResNet50       & {98.51\%}         & {98.90\%}          & {98.10\%}       & {98.50\%}         \\
    \hline
    Weather + SD          & ResNet34       & {90.63\%}         & {88.05\%}          & {94.02\%}       & {90.93\%}         \\
    Weather + SD          & ResNet50       & {95.61\%}         & {95.94\%}          & {95.25\%}       & {95.59\%}         \\
    Comic + SD            & ResNet34       & {92.87\%}         & {88.58\%}          & {98.44\%}       & {93.25\%}         \\
    Comic + SD            & ResNet50       & {98.34\%}         & {100.00\%}         & {96.68\%}       & {98.31\%}         \\
    Small Indoor + DALLE2 & ResNet34       & {51.04\%}         & {80.95\%}          & {2.72\%}        & {5.27\%}          \\
    Small Indoor + DALLE2 & ResNet50       & {51.84\%}         & {82.86\%}          & {4.65\%}        & {8.80\%}          \\
    Weather + DALLE2      & ResNet34       & {50.83\%}         & {53.33\%}          & {13.33\%}       & {21.33\%}         \\
    Weather + DALLE2      & ResNet50       & {50.67\%}         & {56.67\%}          & {5.67\%}        & {10.30\%}         \\
    \hline
  \end{tabular}
  \caption{Precision, Recall and F1-Score on the \% datasets}
  \label{tab:results-on-datasets}
\end{table*}

@@

We experimented with two ResNet models, ResNet34 and XResNet50(ResNet50). We trained both models on our main dataset, which is the Indoor dataset with its generated pictures by Stable Diffusion.  After the models have been trained, we tested them on the following datasets: Weather dataset with its generated pictures by Stable Diffusion (Weather + SD), Comic dataset with its generated pictures by Stable Diffusion (Comic + SD), Small Indoor dataset with its generated pictures by Dalle2 (Small Indoor + DALLE2), and Weather dataset with its generated pictures by Dalle2 (Weather + DALLE2). The result metrics including accuracies, precisions, recalls and F1 scores are shown in Table \ref{tab:results-on-datasets}.

As shown in the table, on the validation set from the main dataset, the ResNet34 model has a higher accuracy, precision, recall and F1 score than the ResNet50 model. However, on the testing datasets, the ResNet50 model generally has a higher accuracy, precision, recall and F1 score than the ResNet34 model in most comparisons except for "Weather + DALLE2" dataset where ResNet50 is worse in F1 score. Even though the all the difference is within 10\%, this is a statistically significant trend that ResNet50 seems to generalize better on the dataset that it hasn't seen. This is likely due to the fact that the ResNet50 model has more layers than the ResNet34 model, which makes it more complex and has more parameters to learn. Thus, it can learn more features from the training dataset and generalize better to the testing datasets. However, the ResNet34 model is simpler and has less parameters to learn. Thus, it is more likely to overfit the training dataset and perform worse on the testing datasets.

\begin{figure}[ht]
  \includegraphics[width=190pt]{./assets/paper_training_confusion_matrix.png}
  \caption{Training Confusion Matrix on Weather}
  \label{fig:paper_training_confusion_matrix}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=190pt]{./assets/paper_dalle2_weather_confusion_matrix.png}
  \caption{Testing Confusion Matrix on Weather + DALLE2}
  \label{fig:paper_dalle2_weather_confusion_matrix}
\end{figure}

The models have shown surprisingly good results with the comic book dataset where both the synthetic images and the original images are not camera-captured photos and also very different from the training dataset in styles and themes. The fact that the models generalize fairly well on this dataset indicates that the model may have learned some general features of the synthetic images that are from the Stable Diffusion model. But the much worse performance from the two models on the DALL·E-2-generated datasets indicates that the features the models have learned are very specific to the Stable Diffusion model and not generalizable to other models, even when those models are also based on diffusers. To human eyes, DALL·E 2 do generate more realistic and detailed images compared to the Stable Diffusion model. From the comparison of the confusion matrices shown in Figure \ref{fig:paper_training_confusion_matrix} and Figure \ref{fig:paper_dalle2_weather_confusion_matrix}, we can see that for the validation split for the main dataset that we trained on, the models predicted rather balanced results. However, the models are more likely to predict the DALL·E-2 generated images as original images thinking that they are real photos.

\section{Conclusion, Discussion and Future Work}

In this article, we contributed:

\begin{itemize}
  \item We generated labeled datasets with original pictures and AI-generated pictures.
  \item We trained models with ResNet and SD-generated pictures, which have a high accuracy on detecting the AI pictures generated by Stable Diffusion.
  \item We found that the pictures generated by Stable Diffusion have significant differences comparing to real pictures. Using the ResNet50 model trained with SD-generated pictures, we have more than $95\%$ accuracy and F1-score on detecting such difference.
  \item However, both of the models generated by ResNet are not able to detect pictures generated by DALLE2, with nearly 50\% accuracy and less than 15\% recall.
\end{itemize}

We think there might be 2 reason for why our model cannot detect any pictures generated by DALLE2. 1) DALLE2 might have a great performance that the DALLE2-generated pictures are very similar to real pictures. 2) DALLE2-generated pictures have different characteristic comparing to the SD-generated pictures.

To verify our result, we can train models with DALLE2-generated pictures and test whether it have a high accuracy on detecting them in the future. This is currently limited by funding, because we didn't get free access to DALLE2.
